---
title: "Materials for Project Workday 2/27"
output: pdf_document
date: "`r Sys.Date()`"
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: blue
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("tidyverse", quietly = T)) install.packages("tidyverse")
library(tidyverse)
CensusData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/compiledCensusData.csv",header=T)
```
## Three Specific Research Questions
-How does educational attainment impact the personal earnings of an individual in the United States?
\linebreak -Does the gender of a person impact their personal earnings in the United States? 
\linebreak -Does a person's health exhibit a correlation with their earnings in the United States?

## Clearly Defined Observations
An observation is defined as an individual surveyed by the US Census with the responses for the 8 following variables: 
\linebreak | Race | RACE |
\linebreak | Highest Education | H_ED |
\linebreak | Gender | SEX |
\linebreak | Total Household Income | TOT_IN |
\linebreak | Total Personal Earnings | PEARNVAL |
\linebreak | Marital Status | MARITL |
\linebreak | Health Status | HEA | 
\linebreak | Previous Year of Residence | RES_DIV |
```{r,include=T}
head(CensusData, n=1)
```

## Response Variable
The response variable is "PEARNVAL," referring to the personal earnings of a person. 
```{r}
hist(CensusData$PEARNVAL)
summary(CensusData$PEARNVAL)
```

## Dataset
Chosen data sets: 2020+2021 Annual Social and Economic Benefits, Collected by the US Census
https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html 
https://www.census.gov/data/datasets/2021/demo/cps/cps-asec-2021.html 
```{r}
head(CensusData, n=10)
```