---
title: "United States Income Data - STAT 3220 Project I" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \usepackage{hyperref}
- \onehalfspacing
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```


```{r, include = FALSE}
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
library(ggplot2)
library(dplyr)
library(ggthemes)
if (!requireNamespace("cowplot", quietly = TRUE)) {
  install.packages("cowplot")
}
library(cowplot)
```

\newpage 

# Pledge
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Scheuer 

\newpage
## Background

\begin{itemize}
  \item Does a higher level of educational attainment generally increase personal earnings income across different states?
  \item Does personal earnings increase with an individual's health?
  \item Do older individuals generally earn more money than younger individuals? 
\end{itemize}
Given the rise in inflation and cost of living, exploring the relationships connected to personal earnings across different states is fundamental to understanding how factors like education, health status, and age, influence income disparities at a regional level. This inquiry is grounded in the longstanding debate within economic and social research regarding the return on investment in education, health, and wellness. By examining these relationships across diverse geographical areas, the analysis can uncover nuanced insights into how local economies, policies, and opportunities shape the economic benefits of educational attainment. This approach tackles crucial societal issues related to our economy by revealing the controllable factors correlated with high earners, providing insights into the social class hierarchy that revolves around income and career.

The U.S. Census is pivotal in shaping the nation’s political and economic landscape; its results directly influence the allocation of federal funds toward schools, hospitals, public works, and roads, determined by breakdowns and totals for practically countless variables. The accurate reporting of this data is crucial to ensure each community receives its fair share based on its specific needs (Bureau, 2021). Census data is also crucial to the political sphere with its use in redrawing a multitude of political boundaries to ensure each district contains roughly equal numbers of people (Mather & Scommegna, 2019).  Its political importance extends even to the U.S. House of Representatives, which bases its apportionment of House seats on Census population data, safeguarding the equity of voting power within the nation (Farley, 2020). Ultimately, the Census is vital in the assurance of a fair and functional democracy, extending deep into American life underscoring the importance of its comprehensive methods. 

## Data Description

The dataset was compiled from multiple reputable sources, primarily focusing on the United States demographic, economic, and educational landscapes as of 2020. The core of this dataset originates from the Annual Social and Economic Supplement (ASEC) survey, conducted by the US Census Bureau. This survey incorporates the basic Current Population Survey (CPS)—a vital source for official government statistics on employment and unemployment—alongside supplemental questions that delve into poverty, geographic mobility/migration, and work experience. To enrich this dataset further, unemployment rate data were incorporated from the US Bureau of Labor Statistics, a key authority on labor market activity and working conditions in the US. Additionally, the dataset includes urban population percentages for each state, derived from the US Census Bureau’s Decennial Census of Population and Housing, providing insights into urban versus rural demographics. Sales tax rates by state, as reported by the Tax Foundation, were also integrated to offer a financial perspective.

In preparing this comprehensive dataset, a merge process was used to align the data by State. Given the ASEC survey's extensive individual-level data, averages for variables such as highest level of education attained, sex, work expenditures, and age were calculated and aggregated on a State basis. To ensure consistency and relevance, the dataset was cleaned to include only individuals aged 18 and above, aligning the education and income data more accurately with the adult population. This curated ASEC data was then merged with state-sorted unemployment rate data from the Bureau of Labor Statistics and sales tax rates from the Tax Foundation, ensuring a cohesive dataset that facilitates multifaceted analyses.

During the data preparation process, several potential issues were considered. The exclusion of individuals under 18 could limit insights into the full spectrum of educational engagement and early income patterns. Furthermore, while merging datasets from various sources enhances the dataset’s depth, it also introduces complexities related to data compatibility and consistency. Nonetheless, the sources of this data—namely the US Census Bureau, the Bureau of Labor Statistics, and the Tax Foundation—are renowned for their reliability and the rigorous methodologies they employ, underpinning the overall trustworthiness of the dataset. This merged dataset, while primarily representing a snapshot of 2020, offers a valuable foundation for exploring socio-economic dynamics across the United States.

## Exploratory Data Analysis
```{r, include=FALSE}
p1<-ggplot(incomeData,aes(x=PEARNVAL)) +
  geom_histogram(bins=10) +
  labs(x="Personal Earnings (USD)",y="Count",title="Distribution of Personal Earnings by State") +
  theme_linedraw() +
  theme(axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        title = element_text(size=15)) +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE))
median_earnings <- median(incomeData$PEARNVAL, na.rm = TRUE)
median_earnings
#BOX PLOT of personal earnings (PEARNVAL) vs highest education level (H_ED)
incomeData$H_ED <- factor(incomeData$H_ED, levels = c("Vocational Associate's", "Academic Associate's", "Bachelor's"))
p2<-ggplot(incomeData, aes(x = H_ED, y = PEARNVAL, fill = H_ED)) +
  geom_boxplot() +
  labs(x = "Educational Attainment", y = "Personal Earnings (USD)", fill = "Educational Attainment", title = "Personal Earnings by Educational Attainment across States") +
  theme(axis.text.x = element_text(size= 10, angle = 20, hjust = 1),
        axis.text.y = element_text(size=10),
        title = element_text(size=14),
        legend.text = element_text(size = 13), 
        legend.title = element_text(size = 13),
        legend.key.size = unit(0.5, "cm")) 


#Scatterplot of personal earnings (PEARNVAL) by age
agecor_coefficient <- cor(incomeData$AGE, incomeData$PEARNVAL, use = "complete.obs")
p3<-ggplot(incomeData, aes(x = AGE, y = PEARNVAL)) +
  geom_point() +
  labs(x = "Age", y = "Personal Earnings", title = "Age vs Personal Earnings") +
  annotate("text", x = Inf, y = Inf, label = paste("r = ", round(agecor_coefficient, 3)),
           hjust = 1.05, vjust = 2, size = 5, color = "black") +
  theme(axis.text.x = element_text(size= 10, angle = 45, hjust = 1),
        axis.text.y = element_text(size=10),
        title = element_text(size=15))

#BOXPLOT of personal earnings (PEARNVAL) by health status (HEA) 

p4<-ggplot(incomeData, aes(x=HEA,y=PEARNVAL,fill=HEA)) + 
  geom_boxplot() + 
  labs(x="Health Status",y="Personal Earnings (USD)",fill="Health Status",title="Personal Earnings by Health Status across States") +
  theme_linedraw() +
  theme(axis.text.x = element_text(size= 10),
        axis.text.y = element_text(size=10),
        title = element_text(size=14),
        legend.text = element_text(size = 13), 
        legend.title = element_text(size = 13),
        legend.key.size = unit(0.5, "cm")) 
```
```{r, echo=FALSE, fig.width=7, fig.height=9, out.width='0.50\\linewidth', out.height='0.95\\textheight', results="asis"}
plot_grid(p1, p2, ncol = 1, nrow = 2)
plot_grid(p3, p4, ncol = 1, nrow = 2)
```
```{r, echo=FALSE, warning=FALSE,results=FALSE}
#For histogram
mean(incomeData$PEARNVAL)
median(incomeData$PEARNVAL)

#Summary Stats for PEARNVAL V HEA BOXPLOT:
incomeData |>
  filter(HEA == "Good Health") |>
  summary()

incomeData |>
  filter(HEA == "Very Good Health") |>
  summary()

#Summary Stats for PEARNVAL V H_ED BOXPLOT:
incomeData |>
  filter(H_ED == "Vocational Associate's") |>
  summary()

incomeData |>
  filter(H_ED == "Academic Associate's") |>
  summary()

incomeData |>
  filter(H_ED == "Bachelor's") |>
  summary()


```
The histogram titled “Distribution of Personal Earnings by State” depicts the distribution of the response, personal earnings. The histogram is approximately symmetric and unimodal with a potential outlier at  ~\$95,000. The computed median of the personal earnings is \$55,946.86 and the computed mean is \$58,472.35. Containment of the mean within the range of possible values for the median suggests minimal skew.  
The “Age vs Personal Earnings” scatterplot shows the relationship between the average age of adult state residents and the average personal earnings of the state. There is no apparent pattern, supported by the low correlation coefficient of 0.103 and a wide spread of data points, suggesting a poor, or non-existent, linear relationship amongst these variables.

The boxplot “Personal Earnings by Educational Attainment across States” shows the three levels of educational attainment in the data as being associated with relatively different personal earnings. The boxplots indicate the levels of educational attainment as having significantly different associations with the response, personal earnings; none of the interquartile ranges overlap and medians are different. It’s necessary to note that an educational attainment of a Bachelor’s degree only has one observation, leading to a potential limitation of analysis. States with an educational attainment of academic associates have a much greater spread than those with vocational associates, indicated by the greater IQR. Two potential outliers in the vocational associate’s category could potentially skew prediction. 

The boxplot “Personal Earnings by Health Status across States” displays the relationship between health status and personal earnings. A rather extreme outlier is present in the very good health category, far exceeding the range of other observations. Four less extreme outliers exist in the good health category. Health status appears to have an association with personal earnings, supported by a higher center of the very good health category over the good health category. Both categories of health status have an IQR not containing the other, supporting the existence of association with the response, personal earnings. 

## Conclusion

The histogram of “Distribution of Personal Earnings by State” depicts the distribution of the response variable, Personal Earnings, which represents the average earnings of individuals in each state/territory. The mean of personal earnings is \$58,472.35 and the median is \$55,946.86 The graph demonstrates a unimodal distribution, with approximate normality. An outlier, corresponding to the observation of Washington D.C., is observed, exhibiting average personal earnings significantly higher than those of our other states at \$95,387.40, contributing to a slight right skew of the data. This discrepancy likely arises from the high concentration of government-affiliated jobs in Washington DC, which generally necessitate a bachelor's degree for employment. Given the histogram's mostly normal distribution, unimodal configuration, and the presence of a singular outlier, the data on personal earnings is well-suited for linear regression analysis.

The scatterplot of “Age vs Personal Earnings” depicts the relationship between personal earnings and the average age of adult residents in a state. From the EDA, a correlation coefficient of -0.103 suggests that age has no significant association with personal earnings. There is an outlier corresponding to Washington D.C., indicated by the large difference in personal earnings relative to the other variables. This is most likely due to the high cost of living in a metropolitan city. Due to the low correlation coefficient and the lack of a distinctive pattern, age would likely not be a significant contributor to the model. 

The boxplot “Personal Earnings by Educational Attainment across States” reveals a significant relationship between level of educational attainment and personal earnings. States with an educational attainment of a vocational associate’s degree have the lowest median at \$54,113, exceeded by states with an educational attainment of an academic associate’s with a median at \$62,231, and far exceeded by the territory with an educational attainment of a Bachelor’s degree with a median at \$95,387. This arrangement of centers, increasing with educational attainment, is consistent with the supposition that a higher level of education is associated with higher personal earnings. Logically, this relationship is reflected by a higher educational attainment expanding an individual's skill set, making them suitable for a wider range of job opportunities, potentially leading to higher earnings. A narrower spread on the vocational Associate’s level suggests less upward mobility in personal earning than with an academic Associate’s, which has a greater spread of its IQR. As the Bachelor’s level only has one observation, it would be necessary to remove this level in a regression analysis as building a model with an under-reported level could greatly increase error. 

The boxplot “Personal Earnings by Health Status across States” suggests a significant relationship between health and personal earnings; a good health status resulted in median personal earning of \$54,810, while a very good health status resulted in median personal earning of \$54,810. The data supports the conclusion that better health leads to greater personal earning. However, before utilization in a  model, the unequal reporting of levels must be considered; a very good health status was only reported by 9 territories, with the remainder reporting a good health status. This under-reporting of this level could lead to error in the model. A stratified sampling method may be employed to evenly sample levels of health status if it is believed to contribute significant error to the model. 

In conclusion, our exploratory data analysis, examining the interconnections among demographic, economic, and educational insights derived from the ASEC survey and supplementary sources, has led to significant discoveries regarding the relationships between educational attainment and health status with personal earnings. Firstly, we observed an association between higher levels of educational attainment and greater personal earnings, affirming the pivotal role of education in enhancing economic prospects. Secondly, our analysis revealed an association between health status and personal earnings, indicating that lesser health is associated with diminished earnings. This underscores the critical importance of health in economic stability and success. Interestingly, we found no significant correlation between age and personal earnings, challenging common assumptions about the relationship between advancing age and income. These insights not only respond to our initial research queries but also highlight the complex factors that influence economic outcomes across the United States. This analysis underscores the value of comprehensive data collection and analysis, as exemplified by the decennial census, in informing policy and ensuring equitable resource distribution and opportunities for all citizens.

\newpage

## Appendix A: Data Dictionary

| **Reference Name**                     | **Variable Name** | **Description**                                                                                                                                                                                                                                                  |
|------------------------------------------------------------------------------------|------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **State by FIPS Code**                 | `STATEFIPS`       | A qualitative measure that identifies the U.S. state (or D.C.) corresponding to the observation by a standardized numeric code. The 51 possible levels are discrete, ranging from 1-56, omitting 3, 7, 14, 43, and 52.                                           |
| **State**                              | `State`           | A qualitative measure that identifies the state corresponding to the observation. The 51 possible levels are names of the 50 U.S. states and the District of Columbia.                                                                                          |
| **Educational Attainment**             | `H_ED`            | A qualitative measure that identifies the average of highest education among adult residents of a given state. The three possible levels include a Vocational Associate's Degree, an Academic Associate's Degree, and a Bachelor's Degree.                        |
| **Majority Sex**                       | `SEX`             | A qualitative measure that identifies the predominant sex among a state's adult residents. Two possible levels, male and female, indicate if the adult population of a state is predominately male or female.                                                   |
| **Health Status**                      | `HEA`             | A qualitative measure that reports the average health status of a state's residents. Two levels, very good health and good health indicate the average health status of a state's residents.                                                                    |
| **Personal Earnings**                  | `PEARNVAL`        | A continuous quantitative measure that reports the average personal earnings of a state's residents, reported in U.S. Dollars. Possible values within the data range from $45096.53 to $95387.40.                                                               |
| **Age**                                | `AGE`             | A continuous quantitative measure that reports the average age of a state's adult residents in years. Values range from 40.83460 to 46.39759.                                                                                                                    |
| **Unemployment Rate**                  | `UNEMP_RATE`      | A continuous quantitative measure of a state's unemployment rate from 2020. Unemployment rate is reported as a percentage; the range of possible values within the data is from 4.2% to 13.5%.                                                                  |
| **Sales Tax Rate**                     | `TAX_RTE`         | A continuous quantitative measure of a state's sales tax. Sales Tax Rate is reported as a numerical figure; the range of possible values within the data is from 0.0% (0% sales tax) to 7.25% (7.25% sales tax).                                                |
| **Percentage of Urban Residents**      | `URB_PER`         | A continuous quantitative measure of a state's proportion of urban residents to nonurban residents. This variable is reported as a percentage; the range of possible values within the data is from 38.7% to 100.0%.                                            |
| **Work Expenses**                      | `WRK_SPND`        | A continuous quantitative measure that identifies the average amount of money spent on work-related expenses among residents of a state, reported in U.S. Dollars. Possible values in the data range from $1101.676 to $1463.411.                                |


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
head(incomeData,n=15)
```


\newpage
## Appendix C: References

### Background

- Bureau, U. C. (2021, November 23). Why we conduct the decennial census of Population and Housing. Census.gov. https://tinyurl.com/5fdyh82c

- Mather, M., & Scommegna, P. (2019, March 15). Why is the U.S. Census so important?. Population Reference Bureau https://www.prb.org/resources/importance-of-u-s-census/ 

- Farley, R. (2020, January 31). The importance of census 2020 and the challenges of getting a complete count. Harvard Data Science Review. https://hdsr.mitpress.mit.edu/pub/rosc6trb/release/3 

### Data


- 2020 Unemployment Rates: U.S. Bureau of Labor Statistics. (2024). Unemployment rates for states. U.S. Bureau of Labor Statistics. https://www.bls.gov/lau/lastrk20.htm 
 
- Urban percentage of the population for states, historical. Urban Percentage of the Population for States, Historical | Iowa Community Indicators Program. (2024.). https://www.icip.iastate.edu/tables/population/urban-pct-states 

- State and local sales tax rates, 2020. Tax Foundation. (2024, February 22). https://taxfoundation.org/data/all/state/2020-sales-taxes/ 

- Bureau, U. C. (2022, October 27). 2020 annual social and economic supplements. Census.gov. https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html 

- ASEC 2020 Public Use Data Dictionary. (2020). https://tinyurl.com/3h8vexva
