---
title: "Insert heading" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
```

\begin{center}

#\includegraphics{/Users/kristavaranyak/Downloads/normaldist.jpeg}
remove the pound sign and insert the file path to the image on your computer in the {}

\end{center}

\newpage 

## Introduction
Given the rise in inflation and cost of living, exploring the relationships connected to personal earnings across different states is fundamental to understanding how factors like education, health status, and age, influence income disparities at a regional level. This inquiry is grounded in the longstanding debate within economic and social research regarding the return on investment in education, health, and wellness. By examining these relationships across diverse geographical areas, the analysis can uncover nuanced insights into how local economies, policies, and opportunities shape the economic benefits of educational attainment. This approach tackles crucial societal issues related to our economy by revealing the controllable factors correlated with high earners, providing insights into the social class hierarchy that revolves around income and career.

The U.S. Census is pivotal in shaping the nation’s political and economic landscape; its results directly influence the allocation of federal funds toward schools, hospitals, public works, and roads, determined by breakdowns and totals for practically countless variables. The accurate reporting of this data is crucial to ensure each community receives its fair share based on its specific needs (Bureau, 2021). Census data is also crucial to the political sphere with its use in redrawing a multitude of political boundaries to ensure each district contains roughly equal numbers of people (Mather & Scommegna, 2019).  Its political importance extends even to the U.S. House of Representatives, which bases its apportionment of House seats on Census population data, safeguarding the equity of voting power within the nation (Farley, 2020). Ultimately, the Census is vital in the assurance of a fair and functional democracy, extending deep into American life underscoring the importance of its comprehensive methods.
## Methods and Analysis

## Results

## Conclusions

\newpage

## Appendix A: Data Dictionary

| **Reference Name**                     | **Variable Name** | **Description**                                                                                                                                                                                                                                                  |
|------------------------------------------------------------------------------------|------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **State by FIPS Code**                 | `STATEFIPS`       | A qualitative measure that identifies the U.S. state (or D.C.) corresponding to the observation by a standardized numeric code. The 51 possible levels are discrete, ranging from 1-56, omitting 3, 7, 14, 43, and 52.                                           |
| **State**                              | `State`           | A qualitative measure that identifies the state corresponding to the observation. The 51 possible levels are names of the 50 U.S. states and the District of Columbia.                                                                                          |
| **Educational Attainment**             | `H_ED`            | A qualitative measure that identifies the average of highest education among adult residents of a given state. The three possible levels include a Vocational Associate's Degree, an Academic Associate's Degree, and a Bachelor's Degree.                        |
| **Majority Sex**                       | `SEX`             | A qualitative measure that identifies the predominant sex among a state's adult residents. Two possible levels, male and female, indicate if the adult population of a state is predominately male or female.                                                   |
| **Health Status**                      | `HEA`             | A qualitative measure that reports the average health status of a state's residents. Two levels, very good health and good health indicate the average health status of a state's residents.                                                                    |
| **Personal Earnings**                  | `PEARNVAL`        | A continuous quantitative measure that reports the average personal earnings of a state's residents, reported in U.S. Dollars. Possible values within the data range from $45096.53 to $95387.40.                                                               |
| **Age**                                | `AGE`             | A continuous quantitative measure that reports the average age of a state's adult residents in years. Values range from 40.83460 to 46.39759.                                                                                                                    |
| **Unemployment Rate**                  | `UNEMP_RATE`      | A continuous quantitative measure of a state's unemployment rate from 2020. Unemployment rate is reported as a percentage; the range of possible values within the data is from 4.2% to 13.5%.                                                                  |
| **Sales Tax Rate**                     | `TAX_RTE`         | A continuous quantitative measure of a state's sales tax. Sales Tax Rate is reported as a numerical figure; the range of possible values within the data is from 0.0% (0% sales tax) to 7.25% (7.25% sales tax).                                                |
| **Percentage of Urban Residents**      | `URB_PER`         | A continuous quantitative measure of a state's proportion of urban residents to nonurban residents. This variable is reported as a percentage; the range of possible values within the data is from 38.7% to 100.0%.                                            |
| **Work Expenses**                      | `WRK_SPND`        | A continuous quantitative measure that identifies the average amount of money spent on work-related expenses among residents of a state, reported in U.S. Dollars. Possible values in the data range from $1101.676 to $1463.411.                                |


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
# Use head() function to output first several rows

```

\newpage
## Appendix C: Tables and Figures
```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries 
## that will not be included in your appendix
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from EDA, assumptions, and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots
```

\newpage
## Appendix D: References

### Background
1. List your background citations here.

### Data
1. List your data citations here

### Supplemental Code and Analysis Help
1. List your references used to learn more about your techniques and coding here