---
title: "Insert heading" 
author: "Group Name"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
```


\newpage 

# Pledege
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1:
- Member 2:
- Member 3:
- Member 4:

\newpage

## Background

## Data Description

## Exploratory Data Analysis

```{r, echo=FALSE}
## Use this setting of Code chunk for any plots you wish to make
## Refer to the Rmarkdown guide for sizing informaiton on your plots
```


```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk to compute any summary statistics that you want to use
## in your conclusions. 
## These will not print in your report
```

## Conclusion

\newpage

## Appendix A: Data Dictionary

|Variable Name|Abbreviated Name|Description|
|:----:|:----:|:----:|
|  |  |    |

\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
# Use head() function to output first several rows

```

\newpage
## Appendix C: References

### Background
1. List your background citations here.

### Data
1. List your data citations here
