---
title: "STAT 3220 Group Project" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \usepackage{hyperref}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```


```{r, include = FALSE}
# Read in Final Version of Data from GitHub
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
```

\newpage 

# Pledege
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Scheuer 

\newpage
## Background

\begin{itemize}
  \item How does a higher level of educational attainment influence total income across different states, controlling for urbanization and tax rates?
  \item What is the impact of being older, of different genders, and having better health status on total income in various levels of urbanization across states?
  \item: Does a higher unemployment rate within a state, after adjusting for urban percentile and tax percentage, predict a lower average total income, and does this effect vary by sex? 
\end{itemize}

The Census is beyond a historical record or constitutional mandate. The decennial census plays a pivotal role in shaping the political and economic landscape of the nation. Through the meticulous surveying of every resident, the census results directly influence the allocation of hundreds of billions of dollars in federal funding to states, counties, and communities. This funding, critical for schools, hospitals, roads, and public works, is determined by population totals and breakdowns by sex, age, race, and other factors, ensuring that each community receives its fair share based on its specific needs (Bureau, 2021). Beyond funding, census data are instrumental in the equitable distribution of resources for over 100 federal programs, including Medicaid, Head Start, and SNAP. The accuracy of these counts ensures that funding is justly allocated, supporting health, education, housing, and infrastructure programs vital for community well-being.

Redistricting and apportionment are other critical areas where census results hold significant sway. Post-census, state and local officials utilize the updated data to redraw congressional, state, and local district boundaries, ensuring that each district contains roughly equal numbers of people (Mather & Scommegna, 2019). This process is fundamental to maintaining the one-person, one-vote rule, safeguarding the equity of voting power across the nation. Additionally, the apportionment of seats in the U.S. House of Representatives is based on state population counts from the census (Farley, 2020). This means no state has a permanent claim to its current number of House seats; rather, these seats are redistributed according to each state's share of the national population. This redistribution can shift political power, with states experiencing population growth, often in the southern and western regions, gaining seats at the expense of those in the Northeast and Midwest. The initial data from the census, therefore, not only impacts the immediate next decade of federal funding, redistricting, and apportionment but also sets the stage for the political dynamics and resource allocation of the future.

In essence, the decennial census results provide a foundation for a fair and functional democracy, guiding crucial decisions on federal funding, redistricting, and apportionment. These activities, in turn, influence every aspect of American life, from local governance and community services to the national political landscape, underscoring the importance of accurate and comprehensive census data.


## Data Description

The dataset was compiled from multiple reputable sources, primarily focusing on the United States demographic, economic, and educational landscapes as of 2020. The core of this dataset originates from the Annual Social and Economic Supplement (ASEC) survey, conducted by the US Census Bureau. This survey incorporates the basic Current Population Survey (CPS)—a vital source for official government statistics on employment and unemployment—alongside supplemental questions that delve into poverty, geographic mobility/migration, and work experience. To enrich this dataset further, unemployment rate data were incorporated from the US Bureau of Labor Statistics, a key authority on labor market activity and working conditions in the US. Additionally, the dataset includes urban population percentages for each state, derived from the US Census Bureau’s Decennial Census of Population and Housing, providing insights into urban versus rural demographics. Sales tax rates by state, as reported by the Tax Foundation, were also integrated to offer a financial perspective.

In preparing this comprehensive dataset, a merge process was used to align the data by State. Given the ASEC survey's extensive individual-level data, averages for variables such as highest level of education attained, sex, total income, and age were calculated and then aggregated on a State basis. To ensure consistency and relevance, the dataset was cleaned to include only individuals aged 18 and above, thereby aligning the education and income data more accurately with the adult population. This curated ASEC data was subsequently merged with state-sorted unemployment rate data from the Bureau of Labor Statistics and sales tax rates from the Tax Foundation, ensuring a cohesive dataset that facilitates multifaceted analyses.

During the data preparation process, several potential issues were considered. The exclusion of individuals under 18 could limit insights into the full spectrum of educational engagement and early income patterns. Furthermore, while merging datasets from various sources enhances the dataset’s depth, it also introduces complexities related to data compatibility and consistency. Nonetheless, the sources of this data—namely the US Census Bureau, the Bureau of Labor Statistics, and the Tax Foundation—are renowned for their reliability and the rigorous methodologies they employ, underpinning the overall trustworthiness of the dataset. This merged dataset, while primarily representing a snapshot of 2020, offers a valuable foundation for exploring socio-economic dynamics across the United States.

## Exploratory Data Analysis

Outline 
- must include histogram of response variable (total income)
- must include at least four graphical representations and the report should address major features of the data.
- You must include at least one summary that relate to each of your research questions in some way
 
Graphs you can possibly include: 
- Histogram of total income  (PEARNVAL)
- Boxplot of total income (PEARNVAL) by highest education level (H_ED) or Scatterplot of total income (PEARNVAL) by highest education level (H_ED)
- Scatterplot of Average Total Income by Gender across Different Urbanization Levels
- Scatterplot of total income (PEARNVAL) by age
- Boxplot or Scatterplot of total income (PEARNVAL) by health status (HEA) 


```{r}
hist(incomeData$PEARNVAL, ylab="Frequency", xlab="Personal Earnings", main="Histogram of Personal Earnings")
```


```{r, echo=FALSE}
dataForPlot<-incomeData[-6]
par(mfrow=c(3,3))
for (i in names(dataForPlot)[3:9]) {
  if (which(names(dataForPlot) == i) == 4) next  # Skip the response variable
  plot(incomeData[,i], incomeData$PEARNVAL,xlab=i,ylab="Personal Earninngs")
}
```


```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk to compute any summary statistics that you want to use
## in your conclusions. 
## These will not print in your report
```

## Conclusion

Outline: 
- interpret the graphical summaries 
- make sure to tie these all to our 3 research questions
- include a numerical summary that corresponds with the graphical summary 
- include context if there are any large discrepancies 
- use the code chunk above to compute any summary statistics that you want to use in your conclusions. 
- include a description as to why your response variable is
suitable for regression. If it is not suitable, you should discuss what you will
need to do to make it suitable


\newpage

## Appendix A: Data Dictionary

| State | State | Desc |
| State by FIPS Code | STATE | Desc |
| Highest Education | H_ED | Desc |
| Gender | SEX | Desc |
| Total Household Income | TOT_IN | Desc | 
| Total Personal Earnings | PEARNVAL | Desc | 
| Health Status | HEA | Desc | 
| Unemployment Rate | UNEMP_RATE | Desc | 
| Percentage of Urban Residents | URB_PER | Desc | 
| Sales Tax Rate | TAX_RTE | Desc | 


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
head(incomeData, n=15)
```

\newpage
## Appendix C: References

### Background
1. List your background citations here.

Bureau, U. C. (2021, November 23). Why we conduct the decennial census of Population and Housing. Census.gov. https://www.census.gov/programs-surveys/decennial-census/about/why.html#:~:text=The%20results%20of%20the%20census,%2C%20roads%2C%20and%20public%20works. 

Mather, M., & Scommegna, P. (2019, March 15). Why is the U.S. Census so important?. Population Reference Bureau https://www.prb.org/resources/importance-of-u-s-census/ 

Farley, R. (2020, January 31). The importance of census 2020 and the challenges of getting a complete count. Harvard Data Science Review. https://hdsr.mitpress.mit.edu/pub/rosc6trb/release/3 

### Data
1. List your data citations here

\begin{itemize}
  \item 2020 Unemployment Rates: https://www.bls.gov/lau/lastrk20.htm
  \item 2010 Urban Percentage of Population:https://www.icip.iastate.edu/tables/population/urban-pct-states
  \item 2020 Tax Data: https://taxfoundation.org/data/all/state/2020-sales-taxes/
  \item Download for 2020: https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html
  \item Codebook for 2020: https://www.census.gov/programs-surveys/cps/datasets/2020/march/ASEC2020ddl_pub_full.pdf 
\end{itemize}