---
title: "STAT 3220 Group Project" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
# Use this space to load in packages
CensusData <- read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/2020CensusDataSet.csv", header = T)

```

EDA: 
```{r}

more_than_50_unique <- sapply(NSDUH_2020, function(column) {
  if(is.numeric(column)) { 
    return(length(unique(column)) > 50) 
  } else {
    return(FALSE)
  }
})
quantitative_vars <- names(more_than_50_unique[more_than_50_unique])
print(quantitative_vars)
```


\newpage 

# Pledege
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Schuerer 
- Member 4:

\newpage

## Background

## Data Description

## Exploratory Data Analysis

```{r, echo=FALSE}
## Use this setting of Code chunk for any plots you wish to make
## Refer to the Rmarkdown guide for sizing informaiton on your plots
```


```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk to compute any summary statistics that you want to use
## in your conclusions. 
## These will not print in your report
```

## Conclusion

\newpage

## Appendix A: Data Dictionary

|Variable Name|Abbreviated Name|Description|
|:----:|:----:|:----:|
|  |  |    |

\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
# Use head() function to output first several rows

```

\newpage
## Appendix C: References

### Background
1. List your background citations here.

### Data
1. List your data citations here
