---
title: "STAT 3220 Group Project" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

Read in Final Version of Data from GitHub
```{r}
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
```

\newpage 

# Pledege
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Scheuer 

\newpage

## Background

## Data Description

## Exploratory Data Analysis

```{r, echo=FALSE}

```


```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk to compute any summary statistics that you want to use
## in your conclusions. 
## These will not print in your report
```

## Conclusion

\newpage

## Appendix A: Data Dictionary

| State | State | Desc |
| State by FIPS Code | STATE | Desc |
| Highest Education | H_ED | Desc |
| Gender | SEX | Desc |
| Total Household Income | TOT_IN | Desc | 
| Total Personal Earnings | PEARNVAL | Desc | 
| Health Status | HEA | Desc | 
| Unemployment Rate | UNEMP_RATE | Desc | 
| Percentage of Urban Residents | URB_PER | Desc | 
| Sales Tax Rate | TAX_RTE | Desc | 


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}

```

\newpage
## Appendix C: References

### Background
1. List your background citations here.

### Data
1. List your data citations here
Codebook for 2020: https://www2.census.gov/programs-surveys/cps/datasets/2020/march/ASEC2020ddl_pub_full.pdf 
Download for 2020: https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html

2020 Unemployment Rates: https://www.bls.gov/lau/lastrk20.htm 

2010 Urban Percentage of Population: 
https://www.icip.iastate.edu/tables/population/urban-pct-states 

2020 Tax Data
https://taxfoundation.org/data/all/state/2020-sales-taxes/ 