---
title: "United States Income Data - STAT 3220 Project I" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \usepackage{hyperref}
- \onehalfspacing
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```


```{r, include = FALSE}
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
library(ggplot2)
library(dplyr)
library(ggthemes)
if (!requireNamespace("cowplot", quietly = TRUE)) {
  install.packages("cowplot")
}
library(cowplot)
```

\newpage 

# Pledge
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Scheuer 

\newpage
## Background

\begin{itemize}
  \item Does a higher level of educational attainment generally increase personal earnings income across different states?
  \item Does personal earnings increase with an individual's health?
  \item Do older individuals generally earn more money than younger individuals? 
\end{itemize}

The Census is beyond a historical record or constitutional mandate. The decennial census plays a pivotal role in shaping the political and economic landscape of the nation. Through the meticulous surveying of every resident, the census results directly influence the allocation of hundreds of billions of dollars in federal funding to states, counties, and communities. This funding, critical for schools, hospitals, roads, and public works, is determined by population totals and breakdowns by sex, age, race, and other factors, ensuring that each community receives its fair share based on its specific needs (Bureau, 2021). Beyond funding, census data are instrumental in the equitable distribution of resources for over 100 federal programs, including Medicaid, Head Start, and SNAP. The accuracy of these counts ensures that funding is justly allocated, supporting health, education, housing, and infrastructure programs vital for community well-being.

Redistricting and apportionment are other critical areas where census results hold significant sway. Post-census, state and local officials utilize the updated data to redraw congressional, state, and local district boundaries, ensuring that each district contains roughly equal numbers of people (Mather & Scommegna, 2019). This process is fundamental to maintaining the one-person, one-vote rule, safeguarding the equity of voting power across the nation. Additionally, the apportionment of seats in the U.S. House of Representatives is based on state population counts from the census (Farley, 2020). This means no state has a permanent claim to its current number of House seats; rather, these seats are redistributed according to each state's share of the national population. This redistribution can shift political power, with states experiencing population growth, often in the southern and western regions, gaining seats at the expense of those in the Northeast and Midwest. The initial data from the census, therefore, not only impacts the immediate next decade of federal funding, redistricting, and apportionment but also sets the stage for the political dynamics and resource allocation of the future.

In essence, the decennial census results provide a foundation for a fair and functional democracy, guiding crucial decisions on federal funding, redistricting, and apportionment. These activities, in turn, influence every aspect of American life, from local governance and community services to the national political landscape, underscoring the importance of accurate and comprehensive census data.

## Data Description

The dataset was compiled from multiple reputable sources, primarily focusing on the United States demographic, economic, and educational landscapes as of 2020. The core of this dataset originates from the Annual Social and Economic Supplement (ASEC) survey, conducted by the US Census Bureau. This survey incorporates the basic Current Population Survey (CPS)—a vital source for official government statistics on employment and unemployment—alongside supplemental questions that delve into poverty, geographic mobility/migration, and work experience. To enrich this dataset further, unemployment rate data were incorporated from the US Bureau of Labor Statistics, a key authority on labor market activity and working conditions in the US. Additionally, the dataset includes urban population percentages for each state, derived from the US Census Bureau’s Decennial Census of Population and Housing, providing insights into urban versus rural demographics. Sales tax rates by state, as reported by the Tax Foundation, were also integrated to offer a financial perspective.

In preparing this comprehensive dataset, a merge process was used to align the data by State. Given the ASEC survey's extensive individual-level data, averages for variables such as highest level of education attained, sex, total income, and age were calculated and then aggregated on a State basis. To ensure consistency and relevance, the dataset was cleaned to include only individuals aged 18 and above, thereby aligning the education and income data more accurately with the adult population. This curated ASEC data was subsequently merged with state-sorted unemployment rate data from the Bureau of Labor Statistics and sales tax rates from the Tax Foundation, ensuring a cohesive dataset that facilitates multifaceted analyses.

During the data preparation process, several potential issues were considered. The exclusion of individuals under 18 could limit insights into the full spectrum of educational engagement and early income patterns. Furthermore, while merging datasets from various sources enhances the dataset’s depth, it also introduces complexities related to data compatibility and consistency. Nonetheless, the sources of this data—namely the US Census Bureau, the Bureau of Labor Statistics, and the Tax Foundation—are renowned for their reliability and the rigorous methodologies they employ, underpinning the overall trustworthiness of the dataset. This merged dataset, while primarily representing a snapshot of 2020, offers a valuable foundation for exploring socio-economic dynamics across the United States.

## Exploratory Data Analysis
```{r, include=FALSE}
p1<-ggplot(incomeData,aes(x=PEARNVAL)) +
  geom_histogram(bins=10) +
  labs(x="Personal Earnings (USD)",y="Count",title="Distribution of Personal Earnings by State") +
  theme_linedraw() +
  theme(axis.text.x = element_text(size=10),
        axis.text.y = element_text(size=10),
        title = element_text(size=15)) +
  scale_x_continuous(labels = function(x) format(x, scientific = FALSE))

#BOX PLOT of personal earnings (PEARNVAL) vs highest education level (H_ED)
incomeData$H_ED <- factor(incomeData$H_ED, levels = c("Vocational Associate's", "Academic Associate's", "Bachelor's"))
p2<-ggplot(incomeData, aes(x = H_ED, y = PEARNVAL, fill = H_ED)) +
  geom_boxplot() +
  labs(x = "Educational Attainment", y = "Personal Earnings (USD)", fill = "Educational Attainment", title = "Personal Earnings by Educational Attainment across States") +
  theme(axis.text.x = element_text(size= 10, angle = 20, hjust = 1),
        axis.text.y = element_text(size=10),
        title = element_text(size=14),
        legend.text = element_text(size = 13), 
        legend.title = element_text(size = 13),
        legend.key.size = unit(0.5, "cm")) 


#Scatterplot of personal earnings (PEARNVAL) by age
agecor_coefficient <- cor(incomeData$AGE, incomeData$PEARNVAL, use = "complete.obs")
p3<-ggplot(incomeData, aes(x = AGE, y = PEARNVAL)) +
  geom_point() +
  labs(x = "Age", y = "Personal Earnings", title = "Age vs Personal Earnings") +
  annotate("text", x = Inf, y = Inf, label = paste("r = ", round(agecor_coefficient, 3)),
           hjust = 1.05, vjust = 2, size = 5, color = "black") +
  theme(axis.text.x = element_text(size= 10, angle = 45, hjust = 1),
        axis.text.y = element_text(size=10),
        title = element_text(size=15))

#BOXPLOT of personal earnings (PEARNVAL) by health status (HEA) 

p4<-ggplot(incomeData, aes(x=HEA,y=PEARNVAL,fill=HEA)) + 
  geom_boxplot() + 
  labs(x="Health Status",y="Personal Earnings (USD)",fill="Health Status",title="Personal Earnings by Health Status across States") +
  theme_linedraw() +
  theme(axis.text.x = element_text(size= 10),
        axis.text.y = element_text(size=10),
        title = element_text(size=14),
        legend.text = element_text(size = 13), 
        legend.title = element_text(size = 13),
        legend.key.size = unit(0.5, "cm")) 
```
```{r, echo=FALSE, fig.width=7, fig.height=9, out.width='0.50\\linewidth', out.height='0.95\\textheight', results="asis"}
plot_grid(p1, p2, ncol = 1, nrow = 2)
plot_grid(p3, p4, ncol = 1, nrow = 2)
```
```{r, echo=FALSE, warning=FALSE,results=FALSE}
#For histogram
mean(incomeData$PEARNVAL)
median(incomeData$PEARNVAL)

#Summary Stats for PEARNVAL V HEA BOXPLOT:
incomeData |>
  filter(HEA == "Good Health") |>
  summary()

incomeData |>
  filter(HEA == "Very Good Health") |>
  summary()

#Summary Stats for PEARNVAL V H_ED BOXPLOT:
incomeData |>
  filter(H_ED == "Vocational Associate's") |>
  summary()

incomeData |>
  filter(H_ED == "Academic Associate's") |>
  summary()

incomeData |>
  filter(H_ED == "Bachelor's") |>
  summary()


```
The histogram titled “Distribution of Personal Earnings by State” depicts the distribution of the response variable, personal earnings. The histogram shows approximate symmetry, unimodality, and a potential outlier at approximately \$95,000. The median of the personal earnings appears to be between ~\$53,000 - ~\$59,000. The computed mean of personal earnings is \$58472.35. Containment of the mean within the range of possible values for the median suggests minimal skew.  
The “Age vs Personal Earnings” scatterplot shows the relationship between the average age of adult state residents and the average personal earnings of the state. There is no apparent pattern, supported by the low correlation coefficient of 0.103 and a wide spread of data points, suggesting a poor, if any, linear relationship amongst these variables.
The boxplot “Personal Earnings by Educational Attainment across States” shows the three levels of educational attainment levels in the data as being associated with relatively different personal earnings. States with an educational attainment of a vocational associate’s degree have the lowest median at \$54,113, exceeded by states with an educational attainment of an academic associate’s with a median at \$62,231, and far exceeded by the territory with an educational attainment of a Bachelor’s degree with a median at \$95,387. The boxplots indicate the levels of educational attainment as being significantly different when considered with the response, personal earnings; none of the interquartile ranges overlap. It’s necessary to note that an educational attainment of bachelor’s degree only has one observation, leading to a potential limitation of analysis. States with an educational attainment of academic associates have a much greater spread than those with vocational associates, indicated by the greater IQR. Two potential outliers exist in the vocational associate’s category, which could potentially skew prediction. 
The boxplot “Personal Earnings by Health Status across States” displays the relationship between health status and personal earnings. A rather extreme outlier is present in the very good health category, far exceeding the range of other observations. Four less extreme outliers exist in the good health category. Health status appears to have an association with personal earnings, supported by a higher center of the very good health category (median \$65,425) over the good health category, centered with a median at \$54,810. Both categories of health status have an IQR not containing the other, supporting the existence of association with the response, personal earnings. 

## Conclusion

The histogram of “Distribution of Personal Earnings by State” depicts the distribution of the response variable, Personal Earnings, which represents the average earnings of individuals in each state/territory. The mean of personal earnings is \$58472.35 and the median is \$55946.86 The graph demonstrates a unimodal distribution, with approximate normality. An outlier, corresponding to the observation of Washington D.C., is observed, exhibiting average personal earnings significantly higher than those of our other states at \$95387.40, contributing to a slight right skew of the data. This discrepancy likely arises from the high concentration of government-affiliated jobs in Washington DC, which generally necessitate a bachelor's degree for employment. Given the histogram's mostly normal distribution, unimodal configuration, and the presence of a singular outlier, the data on personal earnings is well-suited for linear regression analysis
\\\\ AGE v. PEARNVAL SCATTER
\\\\ H_ED v. PEARNVAL BOXPLOT
\\\\ HEA v. PEARNVAL BOXPLOT
In conclusion, our exploratory data analysis, examining the interconnections among demographic, economic, and educational insights derived from the ASEC survey and supplementary sources, has led to significant discoveries regarding the relationships between educational attainment and health status with personal earnings. Firstly, we observed an association between higher levels of educational attainment and greater personal earnings, affirming the pivotal role of education in enhancing economic prospects. Secondly, our analysis revealed an association between health status and personal earnings, indicating that lesser health is associated with diminished earnings. This underscores the critical importance of health in economic stability and success. Interestingly, we found no significant correlation between age and personal earnings, challenging common assumptions about the relationship between advancing age and income. These insights not only respond to our initial research queries but also highlight the complex factors that influence economic outcomes across the United States. This analysis underscores the value of comprehensive data collection and analysis, as exemplified by the decennial census, in informing policy and ensuring equitable resource distribution and opportunities for all citizens.

\newpage

## Appendix A: Data Dictionary
Variables are reported in the following format:

| Reference Name | Variable Name | 

Description 

--------------------------------------------
| \textbf{State by FIPS Code }| STATEFIPS |

A qualitative measure that identifies the U.S. state (or D.C.) corresponding to the observation by a standardized numeric code. The 51 possible levels are discrete, ranging from 1-56, omitting 3, 7, 14, 43, and 52. 

| \textbf{State} | State |

A qualitative measure that identifies the state corresponding to the observation. The 51 possible levels are names of the 50 U.S. states and the District of Columbia. 

| \textbf{Educational Attainment} | H_ED |

A qualitative measure that identifies the average of highest education among adult residents of a given state. The three possible levels include a Vocational Associate's Degree, an Academic Associate's Degree, and a Bachelor's Degree. 

| \textbf{Majority Sex} | SEX | 

A qualitative measure that identifies the predominant sex among a state's adult residents. Two possible levels, male and female, indicate if the adult population of a state is predominately male or female. 

| \textbf{Health Status} | HEA |

A qualitative measure that reports the average health status of a state's residents. Two levels, very good health and good health indicate the average health status of a state's residents. 

| \textbf{Total Household Income} | TOT_IN |

A qualitative measure that identifies the average total household income of adult residents of a state. 13 possible levels in the data range from \$42,500 to \$44,999 to \$70,000 to \$72,499, each capturing a \$2,499 range in household income. 

| \textbf{Personal Earnings} | PEARNVAL | 

A continuous quantitative that reports the average personal earnings of a state's residents, reported in U.S. Dollars. Possible values within the data range from 45096.53 to 95387.40, indicating \$45096.53 and \$95387.40, respectively. 

| \textbf{Age} | AGE |

A continuous quantitative measure that reports the average age of a state's adult residents in years. Values range from 40.83460 to 46.39759. An age of 40.83460 indicates a person who is 40 years old and approximately 305 days old. 

| \textbf{Unemployment Rate} | UNEMP_RATE | 

A continuous quantitative measure of a state's unemployment rate from 2020. Unemployment rate is reported as a percentage; the range of possible values within the data is from 4.2 (4.2% of the adult aged population of the state is unemployed) to 13.5 (13.5% of the state's adult aged population is unemployed). 

| \textbf{Sales Tax Rate} | TAX_RTE | 

A continuous quantitative measure of a state's sales tax. Sales Tax Rate is reported as a numerical figure; the range of possible values within the data is from 0.0 (0% sales tax) to 0.07250 (7.25% sales tax). 

| \textbf{Percentage of Urban Residents} | URB_PER |

A continuous quantitative measure of a state's proportion of urban residents to nonurban residents. This variable is reported as a percentage; the range of possible values within the data is from 38.7 (38.7% of the state's population lives in an urban area) to 100.0 (the entirety of the state's population lives in urban areas). 


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
head(incomeData,n=15)
```


\newpage
## Appendix C: References

### Background

Bureau, U. C. (2021, November 23). Why we conduct the decennial census of Population and Housing. Census.gov. https://tinyurl.com/5fdyh82c
Mather, M., & Scommegna, P. (2019, March 15). Why is the U.S. Census so important?. Population Reference Bureau https://www.prb.org/resources/importance-of-u-s-census/ 
Farley, R. (2020, January 31). The importance of census 2020 and the challenges of getting a complete count. Harvard Data Science Review. https://hdsr.mitpress.mit.edu/pub/rosc6trb/release/3 

### Data

\begin{itemize}
  \item 2020 Unemployment Rates: https://www.bls.gov/lau/lastrk20.htm
  \item 2010 Urban Percentage of Population: https://www.icip.iastate.edu/tables/population/urban-pct-states
  \item 2020 Tax Data: https://taxfoundation.org/data/all/state/2020-sales-taxes/
  \item Download for 2020: https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html
  \item Codebook for 2020: "https://tinyurl.com/3h8vexva"
\end{itemize}