---
title: "The Highest Earner: Factors that affect personal earnings in the United States" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \usepackage{hyperref}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, include=FALSE}
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
library(dplyr)
library(ggplot2)
library(corrplot)
library(car)
library(caret)
library(cowplot)
incomeData<-incomeData |>
  filter(STATEFIPS != 11)
subsetincomeData<-incomeData[-c(11,19,45,21),]
subsetincomemod<-lm(PEARNVAL~URB_PER+H_ED,data=subsetincomeData)
```


\begin{center}

remove the pound sign and insert the file path to the image on your computer in the {}

\end{center}

\newpage 

## Introduction

\begin{itemize}
  \item Does a higher level of educational attainment generally increase personal earnings income across different states?
  \item Do personal earnings increase with an individual's health?
  \item Do older individuals generally earn more money than younger individuals? 
\end{itemize}

Given the rise in inflation and cost of living, exploring the relationships connected to personal earnings across different states is fundamental to understanding how factors that influence income disparities at a regional level. This inquiry is grounded in the longstanding debate regarding the return on investment in education, health, and wellness. By analyzing U.S. Census data, we can gain insight into the nation's political and economic structures, examining how local economies, policies, and opportunities influence each community. This analysis underscores the importance of ensuring that every community receives its fair share of resources, tailored to its unique needs (Bureau, 2021). Also, it is crucial to the political sphere with its use in redrawing political boundaries to ensure each district contains roughly equal numbers of people, addressing funding disparities (Mather & Scommegna, 2019). 

The dataset was merged on a state basis, focusing on individuals 18 and older to better represent the adult population. It includes averages of education level, gender, work expenses, and age from the ASEC survey, combined with state unemployment rates from the Bureau of Labor Statistics and state sales tax rates from the Tax Foundation.

## Methods and Analysis

Before building the model, the dataset was subsetted to exclude the observation corresponding to the District of Columbia (D.C.). Exploratory data analysis revealed that it was the sole territory where "Bachelor's Degree" was reported as the highest education attainment, H_ED. As H_ED was believed to be a significant predictor in the model, D.C. was excluded from the model to prevent skew or extrapolation. Following the removal of this observation, a histogram of the response, personal earnings, appeared unimodal with moderate right skew and with minimal outliers. The first stage in building the model involved fitting it with quantitative predictors. Quantitative variables were examined for evidence of multicollinearity through correlative plots; no concerning relationships were found, but multicollinearity will be reassessed in the final model. Scatterplots of each quantitative variable with the response showed varying degrees of association. Moderate to strong relationships with the response existed with four variables: unemployment rate, tax rate, work expenses, and urban percentage. These four variables were used to build a model, resulting in a globally significant model. The model was reduced via the performance of individual t-tests to include only the two most significant quantitative predictors, urban percentage and work expenses, which had p-values of 0.00879 and 0.01667, respectively. 

Next, qualitative predictors were added to the model. The examination of boxplots of the three qualitative variable with the response suggested that alternative levels of sex does not significantly impact earnings; the median personal earnings of states with a predominantly female workforce is approximately equivalent to those with a predominantly male workforce. However, differing levels of education and health status suggested significant different responses to personal earnings, indicated by non-overlapping interquartile ranges of differing levels for these variables. Therefore, these two quantitative variables, highest education level and health status were added to the model in the second stage of the model building process. The proposed model demonstrated significance by the global f-test. Individual t-tests were then performed to build a model with two predictors: urban percentage and highest educational attainment. The quantitative work expenses predictor became insignificant upon that addition of qualitative predictors, and health status did not demonstrate individual significance. No interactions were believed to be influencing the model, but an interaction between the two remaining main effects within the model was explored. A grouped by scatterplot, plotting PEARNVAL as the response, URB_PER as the quantitative explanatory variable, and H_ED as the qualitative explanatory variable demonstrated no difference in slope when regression lines were plotted through the grouped points. 

From this process, the equation of the proposed model is: $PEARNVAL = 47415.20 + 191.57URB\_PER -6997.66H\_EDvocationalassociates$. An examination of VIFs suggested no concern for multicollinearity with an average and highest individual VIF of 1.103. Analysis of residuals showed a lack of fit with no obvious pattern and mean of zero. However, fanning patterns in fitted values and URB_PER residual plots indicated a lack of constant variance. Slight deviations from normality were observed in the qqplot of the residuals of the model, suggested by deviations from linearity on the tail ends of the distribution. Three transformations, logarithmic, exponential, and square root, were attempt to resolve this deviation; however, they did not resolve the violation of normality. The original model, with an untransformed response, was kept as the model is assumed robust to violations of normality due to its sufficient sample size. The lack of constant variance is noted as a limitation of the model. Examination of outliers was performed and 4 observations were removed on the basis of having excessive leverage and Cook's distance. 

With data subsetted to exclude outliers, a new final model is proposed: $PEARNVAL = 47408.52 + 191.95URB\_PER -7017.40H\_EDvocationalassociates$. After verifying the model was trained on data of sufficient size, the external validation technique of data splitting was used to examine the model for overfitting. The data was split randomly into two subsets of equal size, one used to estimate model parameters and the other used to assess the model's predictive ability. 

## Results

$PEARNVAL = 47408.52 + 191.95URB\_PER -7017.40H\_EDvocationalassociates$

This model is globally significant with a p-value < 0.0001 and an adjusted R² value of 0.56, indicating that this model accounts for 56% of the variation in the data. Additionally, The R² value of 0.5792 suggests that approximately 57.92% of the variance in personal earnings can be explained by urban percentile and higher education. The model 

Cross-splitting produced a model with an R²=0.526, RMSE=4814.87, and MAE=3589.583. 

## Conclusions
Our model shows that increase in urban percentile (URB_PER) leads to higher personal earnings whereas a lower level of educational attainment (H\_EDvocationalassociates) results in lower personal earnings in the United States. With a low R² of 0.5792, the model is not a strong predictor of personal earnings.

Further, data from cross-splitting analysis suggests the model may be slightly overfitted to the training data as the R² value from the model made by cross-splitting analysis is lower than the proposed model. Practically, this means that the proposed model would likely introduce greater error if used for prediction with data external to the training set. 

When estimating the personal earnings of Virginia given an URB_PER of 75.5 and H_EDVassoc of 0, the model predicted a personal earnings of \$61,900.74; the actual earnings were \$71,818.63. The prediction was off by \$9,917.89, a percent error of 13.8%. 

While urban percentage and educational attainment do influence personal income, they may not be the most robust predictors. Additional variables or more complex models could be necessary to capture the full dynamics of affecting personal income. Further research could include other variables not analyzed by our data, such as family size or occupation. Additionally, since we were using 2020 Census data, that could cause a skew towards high unemployment and decrease in personal earnings due to the state of the economy during the global pandemic. Thus, increasing our dataset to beyond 2020 can be beneficial at creating more observations. We could also modify existing variables by treating gender as a continuous proportion instead of a binomial variable. In all, this model is a good starting point for understanding the relationship between urban environment, educational attainment, and personal earnings, but further research is needed to create a stronger model.

\newpage

## Appendix A: Data Dictionary

| **Reference Name**                     | **Variable Name** | **Description**                                                                                                                                                                                                                                                  |
|------------------------------------------------------------------------------------|------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **State by FIPS Code**                 | `STATEFIPS`       | A qualitative measure that identifies the U.S. state (or D.C.) corresponding to the observation by a standardized numeric code. The 51 possible levels are discrete, ranging from 1-56, omitting 3, 7, 14, 43, and 52.                                           |
| **State**                              | `State`           | A qualitative measure that identifies the state corresponding to the observation. The 51 possible levels are names of the 50 U.S. states and the District of Columbia.                                                                                          |
| **Educational Attainment**             | `H_ED`            | A qualitative measure that identifies the average of highest education among adult residents of a given state. The three possible levels include a Vocational Associate's Degree, an Academic Associate's Degree, and a Bachelor's Degree.                        |
| **Majority Sex**                       | `SEX`             | A qualitative measure that identifies the predominant sex among a state's adult residents. Two possible levels, male and female, indicate if the adult population of a state is predominately male or female.                                                   |
| **Health Status**                      | `HEA`             | A qualitative measure that reports the average health status of a state's residents. Two levels, very good health and good health indicate the average health status of a state's residents.                                                                    |
| **Personal Earnings**                  | `PEARNVAL`        | A continuous quantitative measure that reports the average personal earnings of a state's residents, reported in U.S. Dollars. Possible values within the data range from \$45096.53 to \$95387.40.                                                               |
| **Age**                                | `AGE`             | A continuous quantitative measure that reports the average age of a state's adult residents in years. Values range from 40.83460 to 46.39759.                                                                                                                    |
| **Unemployment Rate**                  | `UNEMP_RATE`      | A continuous quantitative measure of a state's unemployment rate from 2020. Unemployment rate is reported as a percentage; the range of possible values within the data is from 4.2% to 13.5%.                                                                  |
| **Sales Tax Rate**                     | `TAX_RTE`         | A continuous quantitative measure of a state's sales tax. Sales Tax Rate is reported as a numerical figure; the range of possible values within the data is from 0.0% (0% sales tax) to 7.25% (7.25% sales tax).                                                |
| **Percentage of Urban Residents**      | `URB_PER`         | A continuous quantitative measure of a state's proportion of urban residents to nonurban residents. This variable is reported as a percentage; the range of possible values within the data is from 38.7% to 100.0%.                                            |
| **Work Expenses**                      | `WRK_SPND`        | A continuous quantitative measure that identifies the average amount of money spent on work-related expenses among residents of a state, reported in U.S. Dollars. Possible values in the data range from \$1101.676 to \$1463.411.                                |

\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
# Use head() function to output first several rows
head(subsetincomeData,n=15)
```

## Appendix C: Tables and Figures
```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries 
## that will not be included in your appendix
## Figures and Tables Final model output and plots. This should only be a
## summary of your final model. This is limited to 3 pages.
incomemodfinal<-lm(PEARNVAL~URB_PER+H_ED,data=incomeData)
vif(incomemodfinal)
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from EDA, assumptions, and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

#Initial concern for multicollinearity: 
quantincomeData<-incomeData[7:11]
par(mfrow = c(2, 3))
corrplot(cor(quantincomeData))

##EDA to explore variables in relation to the response, PEARNVAL
#Scatter plots for quantitative:

plot(incomeData$AGE,incomeData$PEARNVAL,xlab="AGE",ylab="PEARNVAL")
plot(incomeData$UNEMP_RATE,incomeData$PEARNVAL,xlab="UNEMP_RATE",ylab="PEARNVAL")
plot(incomeData$TAX_RTE,incomeData$PEARNVAL,xlab="TAX_RTE",ylab="PEARNVAL")
plot(incomeData$URB_PER,incomeData$PEARNVAL,xlab="URB_PER",ylab="PEARNVAL")
plot(incomeData$WRK_SPND,incomeData$PEARNVAL,xlab="WRK_SPND",ylab="PEARNVAL")
#Boxplots for qualitative:
par(mfrow=c(2,2))
boxplot(PEARNVAL~H_ED,data=incomeData)
boxplot(PEARNVAL~SEX,data=incomeData)
boxplot(PEARNVAL~HEA,data=incomeData)
#Interactions? 
plot(incomeData$URB_PER,incomeData$PEARNVAL, col=as.factor(incomeData$H_ED))
```

```{r, echo=FALSE,out.width="50%",out.height="50%"} 
# Final model: Assumptions
summary(subsetincomemod)

```

```{r, echo=FALSE, out.width="50%",out.height="50%"}

par(mfrow=(c(2,3)))
plot(incomemodfinal,which=1)
#Normality: 
plot(incomemodfinal,which=2)
hist(residuals(incomemodfinal))
#Lack of Fit & Constant Variance: 
residualPlots(incomemodfinal,tests=F)
```

```{r, echo=FALSE, out.width="50%",out.height="50%"}
#Continued because I needed to mess with the formatting of the rest of the stuff that was in the above chunk.

```


```{r,echo=FALSE}
#outliers
influencePlot(incomemodfinal)
#External Validation Technique: Cross-splitting: 
##Sufficient size assumption met: 
#Yes: n=46>2k+25
#     n=46>29
#Randomly split the data into exact number of cases: 
v<-as.vector(c(rep(TRUE,23),rep(FALSE,23)))
select<-sample(v)
dataforestimation<-subsetincomeData[select,]
dataforprediction<-subsetincomeData[!select,]
### Training
model.cva <- lm(PEARNVAL~URB_PER+H_ED, data=dataforestimation)
### Testing
predictions.cva <- model.cva %>% predict(dataforprediction)
### Evaluating

data.frame(RMSE = RMSE(predictions.cva, dataforprediction$PEARNVAL),
           R2 = R2(predictions.cva, dataforprediction$PEARNVAL),
           MAE = MAE(predictions.cva, dataforprediction$PEARNVAL))
```


\newpage
## Appendix D: References

### Background
- Bureau, U. C. (2021, November 23). Why we conduct the decennial census of Population and Housing. Census.gov. https://tinyurl.com/5fdyh82c

- Mather, M., & Scommegna, P. (2019, March 15). Why is the U.S. Census so important?. Population Reference Bureau https://www.prb.org/resources/importance-of-u-s-census/ 

- Farley, R. (2020, January 31). The importance of census 2020 and the challenges of getting a complete count. Harvard Data Science Review. https://hdsr.mitpress.mit.edu/pub/rosc6trb/release/3 

### Data
- 2020 Unemployment Rates: U.S. Bureau of Labor Statistics. (2024). Unemployment rates for states. U.S. Bureau of Labor Statistics. https://www.bls.gov/lau/lastrk20.htm 
 
- Urban percentage of the population for states, historical. Urban Percentage of the Population for States, Historical | Iowa Community Indicators Program. (2024.). https://www.icip.iastate.edu/tables/population/urban-pct-states 

- State and local sales tax rates, 2020. Tax Foundation. (2024, February 22). https://taxfoundation.org/data/all/state/2020-sales-taxes/ 

- Bureau, U. C. (2022, October 27). 2020 annual social and economic supplements. Census.gov. https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html 

- ASEC 2020 Public Use Data Dictionary. (2020). https://tinyurl.com/3h8vexva

### Supplemental Code and Analysis Help
- https://rpubs.com/muxicheng/1004550