---
title: "Insert heading" 
author: "Group Name"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
```

\begin{center}

#\includegraphics{/Users/kristavaranyak/Downloads/normaldist.jpeg}
remove the pound sign and insert the file path to the image on your computer in the {}

\end{center}

\newpage 

## Introduction

## Methods and Analysis

## Results

## Conclusions

\newpage

## Appendix A: Data Dictionary

|Variable Name|Abbreviated Name|Description|
|:----:|:----:|:----:|
| Variable 1 Full Name |Abbreviated1  | Write the description of the variable  |
| Variable 2 Full Name |Abbreviated2  | Write the description of the variable  |

\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
# Use head() function to output first several rows

```

\newpage
## Appendix C: Tables and Figures
```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries 
## that will not be included in your appendix
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from EDA, assumptions, and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots
```

\newpage
## Appendix D: References

### Background
1. List your background citations here.

### Data
1. List your data citations here

### Supplemental Code and Analysis Help
1. List your references used to learn more about your techniques and coding here