---
title: "STAT 3220 Group Project" 
author: "KHL"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \usepackage{ulem}
- \usepackage{hyperref}
- \onehalfspacing
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```


```{r, include = FALSE}
incomeData<-read.csv("https://raw.githubusercontent.com/kimberlyyliuu/STAT3220Project/main/completeData")
library(ggplot2)
library(dplyr)
library(ggthemes)
```

\newpage 

# Pledege
Please type your names in the appropriate space below. Failing to do so will result in a 0 on this assignment.

"We have neither given nor received unauthorized help on this assignment"

- Member 1: Kimberly Liu
- Member 2: Hadley McQuerrey
- Member 3: Luke Scheuer 

\newpage
## Background

\begin{itemize}
  \item Does a higher level of educational attainment generally increase personal earnings income across different states?
  \item Does personal earnings increase with an individual's health?
  \item Do older individuals generally earn more money than younger individuals? 
\end{itemize}

The Census is beyond a historical record or constitutional mandate. The decennial census plays a pivotal role in shaping the political and economic landscape of the nation. Through the meticulous surveying of every resident, the census results directly influence the allocation of hundreds of billions of dollars in federal funding to states, counties, and communities. This funding, critical for schools, hospitals, roads, and public works, is determined by population totals and breakdowns by sex, age, race, and other factors, ensuring that each community receives its fair share based on its specific needs (Bureau, 2021). Beyond funding, census data are instrumental in the equitable distribution of resources for over 100 federal programs, including Medicaid, Head Start, and SNAP. The accuracy of these counts ensures that funding is justly allocated, supporting health, education, housing, and infrastructure programs vital for community well-being.

Redistricting and apportionment are other critical areas where census results hold significant sway. Post-census, state and local officials utilize the updated data to redraw congressional, state, and local district boundaries, ensuring that each district contains roughly equal numbers of people (Mather & Scommegna, 2019). This process is fundamental to maintaining the one-person, one-vote rule, safeguarding the equity of voting power across the nation. Additionally, the apportionment of seats in the U.S. House of Representatives is based on state population counts from the census (Farley, 2020). This means no state has a permanent claim to its current number of House seats; rather, these seats are redistributed according to each state's share of the national population. This redistribution can shift political power, with states experiencing population growth, often in the southern and western regions, gaining seats at the expense of those in the Northeast and Midwest. The initial data from the census, therefore, not only impacts the immediate next decade of federal funding, redistricting, and apportionment but also sets the stage for the political dynamics and resource allocation of the future.

In essence, the decennial census results provide a foundation for a fair and functional democracy, guiding crucial decisions on federal funding, redistricting, and apportionment. These activities, in turn, influence every aspect of American life, from local governance and community services to the national political landscape, underscoring the importance of accurate and comprehensive census data.


## Data Description

The dataset was compiled from multiple reputable sources, primarily focusing on the United States demographic, economic, and educational landscapes as of 2020. The core of this dataset originates from the Annual Social and Economic Supplement (ASEC) survey, conducted by the US Census Bureau. This survey incorporates the basic Current Population Survey (CPS)—a vital source for official government statistics on employment and unemployment—alongside supplemental questions that delve into poverty, geographic mobility/migration, and work experience. To enrich this dataset further, unemployment rate data were incorporated from the US Bureau of Labor Statistics, a key authority on labor market activity and working conditions in the US. Additionally, the dataset includes urban population percentages for each state, derived from the US Census Bureau’s Decennial Census of Population and Housing, providing insights into urban versus rural demographics. Sales tax rates by state, as reported by the Tax Foundation, were also integrated to offer a financial perspective.

In preparing this comprehensive dataset, a merge process was used to align the data by State. Given the ASEC survey's extensive individual-level data, averages for variables such as highest level of education attained, sex, total income, and age were calculated and then aggregated on a State basis. To ensure consistency and relevance, the dataset was cleaned to include only individuals aged 18 and above, thereby aligning the education and income data more accurately with the adult population. This curated ASEC data was subsequently merged with state-sorted unemployment rate data from the Bureau of Labor Statistics and sales tax rates from the Tax Foundation, ensuring a cohesive dataset that facilitates multifaceted analyses.

During the data preparation process, several potential issues were considered. The exclusion of individuals under 18 could limit insights into the full spectrum of educational engagement and early income patterns. Furthermore, while merging datasets from various sources enhances the dataset’s depth, it also introduces complexities related to data compatibility and consistency. Nonetheless, the sources of this data—namely the US Census Bureau, the Bureau of Labor Statistics, and the Tax Foundation—are renowned for their reliability and the rigorous methodologies they employ, underpinning the overall trustworthiness of the dataset. This merged dataset, while primarily representing a snapshot of 2020, offers a valuable foundation for exploring socio-economic dynamics across the United States.

## Exploratory Data Analysis

Histogram of PEARNVAL
```{r,include=F}
ggplot(incomeData,aes(x=PEARNVAL)) +
  geom_histogram(bins=20) +
  labs(x="Personal Earnings (USD)",y="Count",title="Distribution of Personal Earnings by State") +
  theme_linedraw()

#Scatterplot of personal income (PEARNVAL) vs highest education level (H_ED)
incomeData$Education_Level <- cut(incomeData$H_ED,
                            breaks = c(39, 40, 41, 42, 43),
                            labels = c("Some College but No Degree",
                                       "Associate Degree - Occupational/Vocational",
                                       "Associate Degree - Academic Program",
                                       "Bachelor's Degree"))

ggplot(incomeData, aes(x = H_ED, y = PEARNVAL, color = Education_Level)) +
  geom_point() +
  scale_color_manual(values = c("High School Graduate" = "green",
                                "Some College but No Degree" = "lightpink",
                                "Associate Degree - Occupational/Vocational" = "blue",
                                "Associate Degree - Academic Program" = "orange",
                                "Bachelor's Degree" = "black")) +
  labs(x = "Highest Education (Numeric)", y = "Personal Earnings (USD)", color = "Educational Attainment", title = "Educational Attainment vs Personal Earnings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Scatterplot of total income (PEARNVAL) by age
agecor_coefficient <- cor(incomeData$AGE, incomeData$PEARNVAL, use = "complete.obs")
ggplot(incomeData, aes(x = AGE, y = PEARNVAL)) +
  geom_point() +
  labs(x = "Age", y = "Personal Earnings", title = "Age vs Personal Earnings") +
  annotate("text", x = Inf, y = Inf, label = paste("r = ", round(agecor_coefficient, 3)),
           hjust = 1.05, vjust = 2, size = 5, color = "black") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

#Scatterplot of total income (PEARNVAL) by health status (HEA) 
incomeData$Health_Status<-cut(incomeData$HEA,
                              breaks= c(0,1,2,3,4,5),
                              labels= c("Excellent Health",
                                        "Very Good Health",
                                        "Good Health",
                                        "Fair Health",
                                        "Poor Health"))
heacor_coefficient<-cor(incomeData$HEA,incomeData$PEARNVAL)
ggplot(incomeData, aes(x=HEA,y=PEARNVAL,color=Health_Status)) + 
  geom_point() + 
  scale_color_manual(values=c("Excellent Health" = "purple",
                              "Very Good Health" = "blue",
                              "Good Health" = "orange",
                              "Fair Health" = "black",
                              "Poor Health" = "deepskyblue4")) +
  labs(x="Health Status (Numeric)",y="Personal Earnings (USD)",color="Health Status",title="Health Status vs Personal Earnings") +
  theme_linedraw() +
  annotate("text", x=Inf, y=Inf, label=paste("r = ", round(heacor_coefficient,3)), 
           hjust=1.1, vjust=2, size=5, color="black")
```

The histogram titled Distribution of Personal Earnings by State depicts the distribution of the response variable PEARNVAL which represents the average earnings of individuals in each state/territory. The graph is unimodal, mostly normally distributed, and displays minimal right skew. Because of this, personal earnings are appropriate for linear regression analysis.

	The scatterplot Education Attainment vs Personal Earnings displays the relationship between a state's average education and the average personal earnings of its citizens. The graph has a linear slope with a correlation coefficient of .8055 which suggests a strong relationship between the response and explanatory variable. There is one outlier, Washington DC, whose highest education and personal earnings is significantly higher than the rest.
	
	The Age vs Personal Earnings scatterplot shows the relationship between the average age of a state(excluding individuals under the age of 18) and the average personal earnings of the state. The data points seem to not display a pattern and the correlation coefficient of .1031 suggests that their is no linear relationship between the variables.

  The Health Status vs Personal Earnings scatterplot seems to depict a relatively strong negative linear correlation and a correlation coefficient of -0.764 suggests this to be true. Again, Washington D.C. is an outlier. 

```{r, echo=FALSE, warning=FALSE,results=FALSE}
# Use this chunk to compute any summary statistics that you want to use in your conclusions. These will not print in your report
```

## Conclusion
The first graph, the histogram titled Distribution of Personal Earnings by State, depicts the distribution of the response variable PEARNVAL which represents the average earnings of individuals in each state/territory. The graph is unimodal and displays a relatively strong normal distribution pattern. The histogram is slightly skewed right suggesting that a majority of the states have personal earnings below the mean. There is one outlier in the data set, Washington DC, whose average personal earnings are much greater than any of the other states which could account for some of the skewness. Ultimately, personal earnings is appropriate for linear regression analysis due to the mostly normal distribution and unimodal shape of the histogram. 
Next is the scatterplot depicting the average earnings per state compared to the average highest education level obtained in said state. In this graph, the x-axis contains the average education and is displayed numerically with larger numbers representing a higher education attainment. For example, the number 40 represents some college but no degree while the number 43 represents obtaining a bachelor's degree. The correlation coefficient in this scatterplot is .8055 which suggests a strong relationship between the explanatory and response variable. The scatterplot suggests that, in general, the higher a state's average educational attainment the higher the average personal earnings. Within the graph, there is one outlier, Washington D.C, however, it is not significant as it falls in line with the pattern created by the rest of the data and does not have a drastic impact on the correlation coefficient. The graph shows that states that average Occupational/Vocational Associate degrees commonly have the least personal earnings with Academic Associate degrees typically earning a bit more and Bachelor's degrees earning the most. In addition, the spread of the states with an average education level being associate degree - Occupational/Vocational is denser than the spread of states with an average of Associate Degree-Academic Program or bachelor degree suggesting that lower education attainments result in a smaller possible range of personal earnings. Overall, because of the strong correlation coefficient and the impact data the graph displays, the variable should be included in the model.
The following scatterplot depicts the relationship between average earnings and the average age of individuals in each respective state. The numbers on the x-axis correspond to the average age of its citizens (not including individuals under 18) I.e., 41 represents 41 years old, 42 represents 42 years old, and so on. Visually, there is no form of relationship between these two variables and a correlation coefficient of only .1031 suggests that no linear relationship is present. Ultimately, this means that the average age of the citizens of a state has virtually no relationship to earnings. Washington D.C. is an outlier in this dataset but has no significance as an outlier. Ultimately, due to the low correlation coefficient and the lack of a distinctive pattern, age should not be included in the model.
	After that is a scatterplot showing the relationship between Personal Earnings and the Health Status variables. Health Status is displayed numerically on the x-axis with lower numbers representing better health statuses. In the census dataset where we obtained the data, the health status responses ranged from 1-5 with the number one representing excellent health and the number five representing poor health. This graph depicts a strong negative linear relationship between the variables with the correlation coefficient being -0.764 strongly suggesting that health status decreases as personal earnings decrease. Ultimately, because of the strong correlation, Health Status should be included in a model.
Overall, because of the graphs we created and the exploratory data analysis, we have preliminary answers to the research questions. First, it seems that higher educational attainments positively influence income meaning that the higher the state's average education, the higher the average income. For the second question, it seems that health status has a positive influence on income, however, at the moment, it seems that gender and age have no impact making this question ambiguous currently. Finally, even with the graphs and EDA, it is still unclear if impact unemployment has on average total income.


\newpage

## Appendix A: Data Dictionary

| State | State | A categorical measure that identifies the state corresponding to the observation. The 51 possible levels are the 50 U.S. states and the District of Columbia. |
| State by FIPS Code | STATE | A categorical measure that identifies the U.S. state (or D.C.) corresponding to the observation by a standardized numeric code. The 51 possible levels are discrete, ranging from 1-56, omitting 3, 7, 14, 43, and 52. |
| Highest Education | H_ED |  
| Gender | SEX | Desc |
| Total Household Income | TOT_IN | Desc | 
| Total Personal Earnings | PEARNVAL | Desc | 
| Health Status | HEA | A categorical measure that reports the average health of a state's residents. This variable reports the average of state's residents' reportings to 5 possible levels that rank health status, 1 = Excellent, 2 = Very good, 3 = Good, 4 = Fair, and 5 = Poor. A non-integer value would indicate the state's residents are intermediate between the two categories corresponding to the adjacent integers. | 
| Unemployment Rate | UNEMP_RATE | A quantitative measure of a state's unemployment rate from 2020. Unemployment rate is reported as a percentage; the range of possible values is from 0.0 (no unemployed persons in the state) to 100.0 (all persons in the state are unemployed). | 
| Sales Tax Rate | TAX_RTE | A quantitative measure of a state's sales tax. Sales Tax Rate is reported as a numerical figure; the range of possible values is from 0.0 (0% sales tax) to 1.0 (100% sales tax). | 
| Percentage of Urban Residents | URB_PER | A quantitative measure of a state's proportion of urban residents to nonurban residents. This variable is reported as a percentage; the range of possible values is from 0.0 (none of the state's population lives in an urban area) to 100.0 (the entirety of the state's population lives in urban areas). |


\newpage
## Appendix B: Data Rows
```{r, echo=FALSE}
head(incomeData[1:12,],n=15)
```


\newpage
## Appendix C: References

### Background

Bureau, U. C. (2021, November 23). Why we conduct the decennial census of Population and Housing. Census.gov. https://tinyurl.com/5fdyh82c
Mather, M., & Scommegna, P. (2019, March 15). Why is the U.S. Census so important?. Population Reference Bureau https://www.prb.org/resources/importance-of-u-s-census/ 
Farley, R. (2020, January 31). The importance of census 2020 and the challenges of getting a complete count. Harvard Data Science Review. https://hdsr.mitpress.mit.edu/pub/rosc6trb/release/3 

### Data

\begin{itemize}
  \item 2020 Unemployment Rates: https://www.bls.gov/lau/lastrk20.htm
  \item 2010 Urban Percentage of Population: https://www.icip.iastate.edu/tables/population/urban-pct-states
  \item 2020 Tax Data: https://taxfoundation.org/data/all/state/2020-sales-taxes/
  \item Download for 2020: https://www.census.gov/data/datasets/2020/demo/cps/cps-asec-2020.html
  \item Codebook for 2020: "https://tinyurl.com/3h8vexva"
\end{itemize}